---
title: "Downloading GBIF Occurrences for Multiple Species in a Polygon"
author: "Your Name"
date: "`r format(Sys.Date())`"
output: html_document
---

# ============================================================
# Download GBIF occurrence data for specific species in a polygon
# Author: [Your Name]
# Date: [Today's date]
# ============================================================

```{r}


# Load necessary packages

library(rgbif)
library(sf)
library(dplyr)
library(ggplot2)
library(lubridate)


```


# ---- 1. Define species of interest ----


```{r}
species_list <- c(
  "Rana temporaria",     # Common frog
  "Bufo bufo",           # Common toad
  "Triturus cristatus",  # Great crested newt
  "Pelophylax ridibundus" # Marsh frog
)
```



# ---- 2. Define polygon (as WKT string) ----
# Example: a small polygon roughly over part of Belgium

```{r}

##wkt_polygon <- "POLYGON((4.0 50.5, 5.0 50.5, 5.0 51.0, 4.0 51.0, 4.0 50.5))"
```

```{r}
wkt_polygon <- "POLYGON((3.31787109375 49.84648456674507, 3.3837890625 49.390954746871806, 5.6689453125 48.94559445308591, 5.6689453125 49.775588649559154, 5.64697265625 50.395912897009076, 4.6142578125 50.68915014165799, 3.69140625 50.28372357925253, 3.31787109375 49.84648456674507))"
```


# ---- 3. Loop over species and download data ----
```{r}

occ_list <- list()

for (sp in species_list) {
  cat("Downloading data for:", sp, "...\n")
  
  res <- occ_search(
    scientificName = sp,
    geometry = wkt_polygon,
    limit = 5000,
    hasCoordinate = TRUE
  )
  
  if (!is.null(res$data)) {
    occ_list[[sp]] <- res$data
  }
}

# ✅ Combine safely
occ_data <- dplyr::bind_rows(occ_list) %>% distinct()

cat("Total records downloaded:", nrow(occ_data), "\n")


```
# ---- 5. Optional: save to CSV ----

```{r}
write.csv(occ_data, "gbif_occurrences_polygon.csv", row.names = FALSE)
cat("Data saved to: gbif_occurrences_polygon.csv\n")
```

# ---- 6. (Optional) Convert to spatial data ----

```{r}
occ_sf <- st_as_sf(
  occ_data,
  coords = c("decimalLongitude", "decimalLatitude"),
  crs = 4326,
  remove = FALSE
)
```


```{r}
plot(st_as_sfc(wkt_polygon), col = NA, border = "blue", main = "GBIF occurrences")
plot(occ_sf["species"], add = TRUE, pch = 20, cex = 0.6, col = "darkgreen")

```
# Clean and prepare dates
```{r}

occ_data <- occ_data %>%
  mutate(
    year = as.numeric(year),
    year = ifelse(is.na(year) & !is.na(eventDate),
                  year(lubridate::ymd(eventDate)),
                  year),
    year_group = cut(year,
                     breaks = seq(floor(min(year, na.rm = TRUE) / 5) * 5,
                                  ceiling(max(year, na.rm = TRUE) / 5) * 5 + 5,
                                  by = 5),
                     right = FALSE,
                     include.lowest = TRUE)
  )
```
# Summarize counts per species and 5-year interval

```{r}

occ_summary <- occ_data %>%
  filter(!is.na(year_group)) %>%
  count(species, year_group, name = "records")
```

# Show table summary
```{r}

knitr::kable(occ_summary, col.names = c("Species", "5-year interval", "Number of Records"))
```

## Bar plot: total records per species across time
```{r}

ggplot(occ_summary, aes(x = year_group, y = records, fill = species)) +
  geom_col(position = "dodge") +
  labs(
    title = "GBIF Occurrences per Species (5-year intervals)",
    x = "5-year interval",
    y = "Number of records",
    fill = "Species"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
## Line plot: trends per species
```{r}

ggplot(occ_summary, aes(x = year_group, y = records, group = species, color = species)) +
  geom_line(linewidth = 1) +
  geom_point() +
  labs(
    title = "Temporal trends of GBIF occurrences per species",
    x = "5-year interval",
    y = "Number of records"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

